---
title: "seminar_4_stan"
author: "esm"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Stan multilevel modelling

Today we will be working on expanding our models to multi-level model, also known as hierarchical models.

These are very useful for dealing with data which has some grouping structure; e.g., we have data multiple individuals, which we assume to both have some similarities, but also some differences. Modelling these types of data as multi-level allows us to account for this.

To work up to this, we will first try out different approaches to dealing with grouped data.

These include:

1.  Complete pooling, i.e., ignore grouping structure and fit as a single unit. We only get group-level parameter estimates, but is cheap and easy. however, our estimates might not actually describe any individuals in our group. Also, we get no estimate of with-in individual variance, or between individual differences.
2.  No pooling, i.e., fit the model to each individual, independently of each other. This can work well for large number of trials, but makes it difficult to summarise parameters at the group level correctly without throwing away uncertainty in parameter estimates. In cases with unbalanced number of trials, this becomes even more challenging.
3.  Partial pooling, i.e., fit a multi-level model, which gives us the best of both worlds; we model the parameters of individuals as belonging to a distribution, which thus directly summarises the parameter at the group level. We set priors on these group-level parameters, and the model "learns" the priors to use for the individual level parameters. Computionally costly, as the parameter count can quickly grow large (also somewhat true for many individual fits, though less so).

```{r}
library(cmdstanr)
library(tidyverse)
library(posterior)
library(tidybayes)
library(bayesplot)
library(patchwork)
```

We will first generate some linear correlated data, with a grouping structure:

```{r}
set.seed(42)
k <- 15 # number of groups/individuals
k_trials <- sample(5:30, size = k, replace = TRUE) # number of obs for each individual (might vary in the real world)
n <- k_trials %>% sum() # total number of trials

slope_mu <- 1 # mean slope for individerne i gruppen
slope_sigma <- 1 # distribution of slopes i gruppen

int_mu <- .5 # mean intercept for individerne i gruppen
int_sigma <- 3 # distribution of intercepts i gruppen

k_sigma <- 2 # sigma for hvert individ (her identisk for alle, men kunne have været forskellige)

k_slopes <- rnorm(k, mean = slope_mu, sd = slope_sigma) # individers slopes
k_ints <- rnorm(k, mean = int_mu, sd = int_sigma) # individers intercepts
dtrue <- tibble( # de sande værdier for alle individer, som vil være vores referenceværdier
  id = 1:k,
  k_trials = k_trials,
  i = k_ints,
  s = k_slopes
)
# her simulerer vi vores faktiske trial-level data, som består af en række individer, som alle er lidt forskellige
d <- tibble(
  id = rep(1:k, times = k_trials),
  x = rnorm(n, sd = 2),
  y = rep(k_ints, times = k_trials) + rep(k_slopes, times = k_trials) * x + rnorm(n = n, sd = k_sigma)
)

# vi plotter den simulerede data + de kendte linjer som vi vil prøve at fitte senere
d %>% 
  ggplot(aes(x, y, color = factor(id)))+
  geom_point()+
  geom_abline(data = dtrue, aes(intercept = i, slope = s, color = factor(id)))+
  facet_wrap(~id)

# samme plot, bare overlap
d %>% 
  ggplot(aes(x, y, color = factor(id)))+
  geom_point()+
  geom_abline(data = dtrue, aes(intercept = i, slope = s, color = factor(id)))

# vi plotter den simulerede data uden gruppestruktur
d %>% 
  ggplot(aes(x, y))+
  geom_point()+
  geom_abline(intercept = int_mu, slope = slope_mu)

```

Compile models: complete pooling, no pooling, and a simple, and an optimised partial pooling model

```{r}
cp <- cmdstan_model("models/linear_complete_pooling.stan")
np <- cmdstan_model("models/linear_no_pooling.stan")
pp <- cmdstan_model("models/linear_partial_pooling.stan")
# ppo <- cmdstan_model("models/linear_partial_pooling_optimised.stan")

# plot of some different options of priors on sigma
tibble(
  x = seq(0, 10, by = 0.01),
  lognorm = dnorm(log(x), mean = 0, sd = 1)/x,
  expprior = dexp(x, 1),
  halfnorm = dnorm(x)*2
) %>% 
  ggplot(aes(x))+
  geom_line(aes(y = lognorm, color = "lognorm"))+
  geom_line(aes(y = halfnorm, color = "halfnorm"))+
  geom_line(aes(y = expprior, color = "exp"))+
  xlim(-1, 10)

```

Fit all 4 models

```{r}
cp_f <- cp$sample(
  data = list(
    n = nrow(d),
    y = d$y,
    x = d$x
  ),
  chains = 4,
  parallel_chains = 4,
  refresh = 1000,
  iter_warmup = 1000,
  iter_sampling = 1000,
)

np_f <- np$sample(
  data = list(
    n = nrow(d),
    y = d$y,
    x = d$x,
    k = length(unique(d$id)),
    id = d$id
  ),
  chains = 4,
  parallel_chains = 4,
  refresh = 1000,
  iter_warmup = 1000,
  iter_sampling = 1000
)
ybar <- mean(d$y)
xbar <- mean(d$x)
xsd <- sd(d$x)

pp_f <- pp$sample(
  data = list(
    n = nrow(d),
    y = d$y,
    x = d$x,
    k = length(unique(d$id)),
    id = d$id
  ),
  chains = 4,
  parallel_chains = 4,
  refresh = 1000,
  iter_warmup = 1000,
  iter_sampling = 1000,
)
# 
# ppo_f <- ppo$sample(
#   data = list(
#     n = nrow(d),
#     y = d$y,
#     x = d$x,
#     k = length(unique(d$id)),
#     id = d$id
#   ),
#   chains = 4,
#   parallel_chains = 4,
#   refresh = 1000,
#   iter_warmup = 1000,
#   iter_sampling = 1000,
# )

```

Her plotter vi først vores complete pooling eksempel
Vi antager at alle observationer kommer fra et enkelt individ
```{r}
dplot <- geom_point(data = d, aes(x, y, color = factor(id)), show.legend = F)
truelines <- geom_abline(data = dtrue, aes(intercept = i, slope = s, color = factor(id)), linetype = "dotted")

cp_f$summary()

p1 <- cp_f %>% 
  spread_draws(beta0, beta1, sigma) %>% 
  expand_grid(x = seq(from = min(d$x), to = max(d$x), length.out = 101)) %>% 
  mutate(
    mu = beta0 + beta1 * x,
    pred = rnorm(n(), mean = mu, sd = sigma) # her har jeg tilføjet predictions fra modellen: kan modellen regenerer vores data? Hvis ikke, skal vi måske være skeptiske
  ) %>% 
  ggplot(aes(x = x, y = mu))+
  geom_point(aes(y = pred), color = "black")+
  dplot+
  stat_lineribbon(color = "black",fill = "red", .width = .89, alpha = 1)+
  ggtitle("Complete pooling")
p1


p1 <- cp_f %>% 
  spread_draws(beta0, beta1, sigma) %>% 
  expand_grid(x = seq(from = min(d$x), to = max(d$x), length.out = 101)) %>% 
  mutate(
    mu = beta0 + beta1 * x,
  ) %>% 
  group_by(x) %>% 
  median_qi(mu) %>% 
  ggplot(aes(x = x, y = mu))+
  dplot+
  geom_lineribbon(aes(ymin = .lower, ymax = .upper), color = "black",fill = "black", alpha = .25)+
  ggtitle("Complete pooling")

p1



```

Her plotter vi så vores no pooling model, hvor vi effektivt fitter en model per individ
Da Stan er vectoriset, kan vi praktisk gøre det i en enkelt model, men det er effektivt det samme
```{r}
p2 <- np_f %>% 
  spread_draws(beta0[id], beta1[id]) %>% 
  expand_grid(x = seq(from = min(d$x), to = max(d$x), length.out = 101)) %>% 
  mutate(
    mu = beta0 + beta1 * x
  ) %>% 
  group_by(id, x) %>% 
  median_qi(mu) %>% 
  ggplot(aes(x = x, y = mu, color = factor(id), fill = factor(id)))+
  dplot+
  geom_lineribbon(aes(ymin = .lower, ymax = .upper, color = factor(id)), alpha = .25)+
  ggtitle("No pooling")+
  truelines+
  facet_wrap(~id)
p2

np_f$summary(variables = c("beta0"))
np_f$summary(variables = c("beta1"))
np_f$summary(variables = c("sigma"))
```

Her plotter vi så vores resultater fra vores partial pooling/multilevel model, hvor vi forsøger både at fitte distributionen af individer, samt enkelte individer inden for den distribution.
Først plotter vi estimatet af gruppen, i.e., det gennemsnitlige individ fra vores pool af individer
```{r}
p3 <- pp_f %>% 
  spread_draws(beta0_mu, beta1_mu) %>%
  expand_grid(x = seq(from = min(d$x), to = max(d$x), length.out = 101)) %>% 
  mutate(
    mu = beta0_mu + beta1_mu * x
  ) %>% 
  ggplot(aes(x = x, y = mu))+
  dplot+
  stat_lineribbon(color = "black", fill = "black",.width = .89, alpha = .25)+
  # stat_summary(fun = "mean", geom = "line", linewidth = 1)+
  ggtitle("Partial pooling, group estimates")
p3
```

Herefter plotter ekstrahere vi de relevante parameter samples, som beskriver de enkelte individer fra vores partial pooling/multilevel model.
```{r}
p4 <- pp_f %>% 
  spread_draws(beta0[id], beta1[id]) %>%  
  expand_grid(x = seq(from = min(d$x), to = max(d$x), length.out = 101)) %>% 
  mutate(
    pred = beta0 + beta1 * x
  ) %>% 
  ggplot(aes(x = x, y = pred, color = factor(id), fill = factor(id)))+
  dplot+
  stat_lineribbon(.width = .89, alpha = .25)+
  # stat_summary(fun = "mean", geom = "line", linewidth = 1)+
  ggtitle("Partial pooling, individual estimates")+
  truelines
p4
```


```{r}
p_all <- (p1+p2+p3+p4)+
  plot_layout(guides = "collect")&
  theme_minimal()+
  theme(legend.position = "bottom", legend.direction = "horizontal")
  
p_all
```

```{r}
pp_f %>% 
  gather_draws(beta0[id], beta1[id], ndraws = 100) %>% 
  ggplot(aes(x = id, y = .value, color = .variable))+
  facet_grid(cols = vars(.variable))+
  stat_pointinterval()

pp_f %>% 
  gather_draws(beta0[id], beta1[id]) %>% 
  mode_hdci() %>%
  ggplot() +
  facet_grid(cols = vars(.variable)) +
  geom_pointinterval(aes(x = id, y = .value, ymin = .lower, ymax = .upper))+
  geom_point(data = dtrue %>% rename(beta0 = i, beta1 = s) %>% pivot_longer(c(beta0, beta1), names_to = ".variable", values_to = "val"), aes(x = id+ .3, y = val), color = "red", alpha = .5)


```
