---
title: "get_started_cmdstanr"
author: "esm"
format: html
editor: visual
---

## Install cmdstanr (command stan r)

First, make sure you are in a fresh session: Go to Session -\> Restart R.

cmdstanr is not on cran, so we need to tell R where to install it from:

```{r}
install.packages("cmdstanr", repos = c('https://stan-dev.r-universe.dev', getOption("repos")))
```



After installing, load the package:\

```{r}
library(cmdstanr)
```

to use cmdstanr, we need a C++ compiler. you may already have one, check with the following function:\

```{r}
check_cmdstan_toolchain()
```

if you get: "The C++ toolchain required for CmdStan is setup properly!", then you can skip the C++ compiler installation and go directly to Install cmdstan

Install C++ compiler: If you are on windows, go to: (if you have an old version of rtools, uninstall it, and download the newest version)\
https://cran.r-project.org/bin/windows/Rtools/rtools45/rtools.html\
If you are on Linux or MacOS, go to link and follow the instructions here:\
https://mc-stan.org/docs/cmdstan-guide/installation.html#cpp-toolchain

once you have installed the C++ compiler, you can then install cmdstan:

```{r}
cmdstanr::install_cmdstan(overwrite = TRUE)
```

If you have any issues during the different steps, refer to the guide at:\
<https://mc-stan.org/cmdstanr/articles/cmdstanr.html>\

## Test cmdstanr

Test the installation with an included example:

```{r}
file <- file.path(cmdstan_path(), "examples", "bernoulli", "bernoulli.stan")

mod <- cmdstan_model(file, force_recompile = TRUE, quiet = FALSE)
mod$print()
```
if the example model is printed, everything works!

finally, install some convenience packages for working with the fits afterwards:
```{r}
install.packages("bayesplot", "posterior", "tidybayes")
```
