---
title: "Seminar exercises week 3: Generalized Context Model"
format: html
editor: visual
---

# Seminar exercises week 3: Generalized Context Model

The goal of these exercises is to estimate the MDS and GCM parameters from data, using maximum likelihood like last week.

Aim to complete exercises 1 - 4. If you are fast, add exercises 5, but wait till after the introduction to Stan.

```{r setup}
library(tidyverse)

theme_set(theme_minimal())

```

# Exercise 1

Familiarize yourself with the multidimensional scaling (MDS) approach, as described in Nosofsky (1986).

Exercise 1 is to estimate the MDS parameters for the 16 stimuli used in Nosofsky (1986) using maximum likelihood and Simplex as optimiser. The confusion matrix for subject 1 is provided below (simply copied from the paper).

Think about how the maximum likelihood looks for discrete data\
(counts of discrete categories).

We will estimate the MDS parameters with dimensional attention weights (w_i) and psychology space scaling parameter (c).

```{r exercise-1-data}

nosofky_subj1<-matrix(
  c(
    250, 94, 0, 0, 192, 42, 0, 0, 7, 5, 0, 0, 0, 0, 0, 0,
    50, 239, 90, 3, 50, 102, 49, 4, 3, 4, 0, 0, 0, 0, 0, 0,
    0, 59, 258, 93, 3, 20, 113, 37, 0, 3, 4, 2, 0, 0, 0, 0,
    0, 11, 137, 288, 0, 3, 60, 79, 0, 1, 6, 5, 0, 0, 0, 0,
    29, 16, 0, 0, 341, 75, 7, 0, 122, 27, 2, 0, 6, 1, 0, 0,
    2, 41, 36, 1, 70, 193, 97, 5, 30, 67, 35, 2, 0, 0, 1, 1,
    0, 7, 48, 22, 5, 40, 226, 104, 0, 17, 61, 20, 0, 0, 0, 0,
    0, 1, 26, 86, 0, 10, 110, 271, 0, 2, 39, 55, 0, 0, 2, 0,
    0, 0, 0, 0, 58, 18, 0, 0, 351, 80, 4, 0, 82, 20, 1, 0,
    1, 0, 1, 0, 13, 55, 40, 6, 77, 255, 95, 9, 16, 45, 19, 0,
    0, 0, 2, 1, 1, 11, 73, 64, 1, 58, 213, 113, 0, 10, 52, 8,
    0, 0, 0, 1, 0, 1, 32, 137, 0, 8, 120, 255, 0, 2, 19, 32,
    0, 0, 0, 0, 1, 2, 1, 0, 111, 23, 1, 0, 309, 96, 6, 1,
    0, 0, 0, 0, 1, 1, 4, 0, 31, 95, 51, 1, 49, 246, 96, 3,
    0, 0, 0, 0, 0, 0, 8, 5, 1, 18, 105, 78, 2, 50, 234, 84,
    0, 0, 0, 0, 0, 0, 1, 2, 0, 2, 56, 186, 0, 2, 101, 249
  ), 
  nrow = 16, 
  byrow = TRUE
)



psych_attr <- 1:4

stim_physical <- crossing(
  angle = psych_attr,
  size = psych_attr
) %>% 
  arrange(size, angle)

stim_physical

heatmapdf <- as.data.frame(as.table(nosofky_subj1))
names(heatmapdf) <- c("row", "column", "value")

heatmapdf$row <- as.numeric(heatmapdf$row)
heatmapdf$column <- as.numeric(heatmapdf$column)

heatmapdf %>% 
  ggplot(aes(x = column, y = row, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(
    option = "magma",
    trans = "log1p"
  ) +
  geom_text(aes(label = value), size = 3)+
  geom_vline(xintercept = seq(4, 12, by = 4)+.5, color = "red", size = 1)+
  geom_hline(yintercept = seq(4, 12, by = 4)+.5, color = "red", size = 1)+
  scale_y_reverse() +
  coord_fixed() +
  labs(
    title = "Heatmap of subject 1 responses",
    x = "Responses",
    y = "Stimuli",
    fill = "Count"
  ) +
  theme_minimal()
```

Here you get a bit of help to get started.

First, a helper function to construct the pairwise comparison

Then, the weighted distance function, and, finally, the similarity function.

```{r exercise-1-functions}
make_pairwise_matrices <- function(stim) {
  crossing(
    i = seq_len(nrow(stim)),
    j = seq_len(nrow(stim))
  )  %>% 
    mutate(
      angle_i = stim$angle[i],
      angle_j = stim$angle[j],
      size_i = stim$size[i],
      size_j = stim$size[j]
    )
}

# note: Distance is here simply euclidiean distance, but weighted by w_: subjects may weigh angle or size higher for their comparison.
distance_gcm <- function(angle_i, angle_j, size_i, size_j, w_angle, r = 2) {
  w_size <- 1 - w_angle 

  (
    w_angle * abs(angle_i - angle_j)^r +
      w_size * abs(size_i - size_j)^r
  )^(1 / r)
}

similarity_gcm <- function(distance, c, k = 2) {
  exp(-c * distance^k)
}
```

Here we define the forward function get_preds_mds, which produces a probability of responding j to stimulus i, given a set of parameters. Afterwards, we utilise the prediction function in our objective function: Remember, our goal here is to calculate how likely the current set of parameters are, given the observed data?\
Compared to earlier examples, data is here in the form of a confusion matrix, but the approach is the same.

```{r exercise-1-mds-prediction-function}
get_preds_mds <- function(par, r = 2, k = 2) {
  x_angle <- par[1:4] # unpack angle
  x_size <- par[5:8] # unpack size
  c_par <- par[9] # scaling par
  w_angle <- par[10] # size/angle weighting par (w_size is 1-w_angle)
  b <- c(1, par[11:25]) # 16 total stimuli, 16 response biases. first is fixed to 1, 15 free.

  stim_mds <- crossing(
    angle = x_angle,
    size = x_size
  ) %>%
    arrange(size, angle)

  pairwise <- make_pairwise_matrices(stim_mds) %>%
    mutate(
      distance = distance_gcm(
        angle_i = angle_i,
        angle_j = angle_j,
        size_i = size_i,
        size_j = size_j,
        w_angle = w_angle,
        r = r
      ),
      similarity = similarity_gcm(distance, c_par, k),
      bias = b[j],
      numerator = similarity * bias
    ) %>%
    group_by(i) %>%
    mutate(pred = numerator / sum(numerator)) %>% # produces a valid probability distribution
    ungroup()

  pairwise %>%
    select(i, j, pred) %>% 
    pivot_wider(
      names_from = j,
      values_from = pred
    ) %>%
    select(-i) %>%
    as.matrix() # return a matrix of same shape as our original data
}

objective_fun_mds <- function(par, data, r = 2, k = 2) {
  if (par[9] < 0) {
    return(1e6) # c must be greater than 0
  }

  if (par[10] < 0 | par[10] > 1) {
    return(1e6) # w_angle must be 0 <= w_angle <= 1
  }
  
  if (any(par[11:25] < 0)) {
    return(1e6) # biases should be non-negative
  }

  pred <- get_preds_mds(par, r = r, k = k)

  -sum(data * log(pred))
}
```

# Exercise 2

Compare the ML solution obtained in Exercise 1 with the observed data.

Assess the goodness-of-fit of the estimated model. Do this first qualitatively with a plot of observed vs predicted counts, then quantify your assessment. R\^2 could be an option for this.

# Exercise 3

One of the strengths of Nosofsky’s (1986) approach is to estimate the coordinates of all stimuli in the 2D psychological space (angle × size) using independent data. These estimates are then used fixed during the estimation of GCM parameters.

That is, we can now reuse our fitted values of Subject 1's psychological space when fitting the Generalised Context Model.

First consider, which parameters remain to be estimated?

Implement an iterative ML estimation using Simplex, starting with the first condition of the second part of the experiment described by Nosofsky (1986).

Follow the principles outlined in the MDS fit. The data for parameter estimation is given in Table 3. The relevant data is in the first two columns of that table.

Assess the quality of the fit quantitatively.

```{r exercise-3-data}
nosofky_subj1_cat_dim <- matrix(
  c(
    213, 4,
    253, 1,
    192, 2,
    218, 1,
    185, 57,
    193, 47,
    187, 40,
    162, 36,
    24, 194,
    33, 198,
    31, 190,
    40, 181,
    0, 204,
    0, 235,
    0, 220,
    0, 258
  ),
  nrow = 16,
  byrow = TRUE
)

cat_dim <- matrix(
  cbind(
    c(1, 0, 0, 1, 0, 1, 1, 0, rep(0, 8)),
    c(rep(0, 8), 0, 1, 1, 0, 1, 0, 0, 1)
  ),
  nrow = 16
)

cat_dim
```

# Exercise 4

Repeat Exercise 3, this time with the remaining three category structures investigated in the article.

In addition, compare the parameter estimates of the four conditions/models.

Which parameter is most relevant for comparison?

# Exercise 5

```{r}
library(cmdstanr)
```

Fit the GCM with MCMC using Stan. The Stan file is provided on Absalon/github.

Consider what would be reasonable priors to use.

Compare the ML estimates with the posterior distribution obtained from Stan.

```{r}
stan_gcm <- cmdstan_model("models/gmc.stan") # you will need to define priors in the stan file before compilation
```
