---
title: "Week 5: BTL and EBA exercises"
author: "esm"
format: html
editor: visual
---

## Exercises

The focus this week is to compare different models to make a decision which one provides a better fit. We perform model comparisons. To get started with this, we use data from a rather classic study (Rumelhart & Greeno, 1971), who modelled the preference of meeting different celebrities. We need the eba package which contains the necessary dataset "celebrities".

We will be working with the BTL and EBA models, in Stan.

Note: I have changed the btl and eba models to use probit instead of logit link function. Hopefully this helps a bit with the familiarity of the models, just be aware that they are changed slightly, meaning that the results are on a different (though comparable) scale :) Models can easily be re-specified to logit scale with 1 line change if you prefer.

Attempt to complete up to and including question 4 for the assignment.

```{r}
# install.packages("eba")
library(eba)
library(cmdstanr)
library(tidyverse)
library(posterior)
library(tidybayes)
library(bayesplot)
library(patchwork)
```

## Exercise 1

Familiarise yourself with the data in celebrities. Write a few comments on what it represents.

```{r}
data(celebrities) # load dataset from lba package
celebrities
```

## Exercise 2

I have provided a starting point in terms transforming the data to be compatible with Stan, and have provided a stan implementation of the BLT and EBA models.

-   A. Run the following block, and inspect the pair_data object to see how it maps to the original celebrities matrix.

-   B. Open the btl.stan file, and have a look through the model code; write, in your own words, how the model works as well as you can.

-   C. Next, compile and fit the btl.stan model, and interpret the results.\
    Consider a plot which can summarise your posterior parameter estimates, by plotting these in relation to the original data.\
    Write a few comments regarding your findings; do they seem to fit the original data well? Does the model struggle in any particular area?

```{r}

make_pair_data <- function(choice_mat) {
  I <- nrow(choice_mat)

  item1 <- c()
  item2 <- c()
  y <- c()
  n <- c()

  for (i in 1:(I - 1)) {
    for (j in (i + 1):I) {
      item1 <- c(item1, i)
      item2 <- c(item2, j)

      y_ij <- choice_mat[i, j]
      y_ji <- choice_mat[j, i]

      y <- c(y, y_ij)
      n <- c(n, y_ij + y_ji)
    }
  }

  list(
    I = I,
    P = length(y),
    item1 = item1,
    item2 = item2,
    y = y,
    n = n
  )
}


pair_data <- make_pair_data(celebrities)

pair_data
```

## Exercise 3

In this exercise you will need to fit the EBA model. I have provided a function below to produce the aspects matrix, which defines an assumed grouping structure in the data, here according to whether the celebrity is a politician, athlete or movie star.

-   A. Inspect the X matrix, and create the eba_data object.

-   B. Inspect the eba.stan file, and describe in your own words how it works. Note the similarity to the BTL model.

-   C. Fit the eba.stan model, and interpret the results. Continue on the plot from exercise two, or create a new one, which summarises both the BLT and EBA model in their fit to the data. Write a few comments on the relative fits of both models to the data. Does the EBA appear better in any respects?

```{r}
make_incidence_matrix <- function() {
  X <- diag(9)
  X <- cbind(X, c(rep(1, 3), rep(0, 3), rep(0, 3))) # policiticians
  X <- cbind(X, c(rep(0, 3), rep(1, 3), rep(0, 3))) # athletes
  X <- cbind(X, c(rep(0, 3), rep(0, 3), rep(1, 3))) # movie stars
  return(X)
}

X <- make_incidence_matrix()

eba_data <- c(
  pair_data,
  list(
    A = ncol(X), # number of unique aspects, incl individual and grouping
    X = X
  )
)
```

## Exercise 4

Now that we have plotted and summarised the two models, we wish to do some more formal model comparison. For this we will use PSIS-LOO-CV.

-   A. Go to the stan code of both models, and add a generated quantities block. Add a calculation for the loglikelihood of each observation, for each draw (see sdt_simple_with_gen_quants.stan for example approach).

-   B. Re-compile and refit both models (remember, adding the generated quantities block does not actually affect fitting, only saves additional data).

-   C. Compare both models with LOO. You can use the following code structure for this (will only work once you save loglikelihood for each obs, here called "log_lik").

-   D. What does our our model comparison tell us? What exactly does our LOO-CV comparison show? Write a few comments with your conclusions.

```{r}
btl_fit_loo <- btl_fit$loo(variables = "log_lik", moment_match = TRUE)

eba_fit_loo <- eba_fit$loo(variables = "log_lik", moment_match = TRUE)

# printing the results of each will also tell us if any obs are problematic for loo. If they are, you can add fit$loo(moment_match = TRUE), which better accounts for the outlier data. will take a fair bit longer to run.
btl_fit_loo 
eba_fit_loo
loo_compare(list(EBA = eba_fit_loo, BTL = btl_fit_loo))
```

## Exercise 5 

Write out the likelihood functions for BTL and EBA and fit with optim() - consider the functions we used for MDS as inspiration. Compare results with your Bayesian estimates, and the observed data.

## Exercise 6

EBA can handle items appearing in multiple categories. Consider other categories which raters may have used in the current dataset: nationality, gender, currently active (at time of data collection), etc.

Construct a new coincidence matrix and refit; use LOO-CV to check if your new model is better at predicting unseen data.
